Embedding Dimension of Polyhedral Losses

### 2020-01-30 colt-20/embed-dim.tex
Took a pass on examples section, moved a lot to the Appendix

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: Abstain loss}\label{sec:examples}
\subsection{Abstain, $\alpha \leq 1/2$, $d=1$}\label{subsec:abstain-d1}
One classification-like problem that is of particular interest is the abstain property, elicited by the loss $\ellabs{\alpha}$ given in Equation~\ref{eq:abstain}.
The property $\gamma = \abstain{\alpha}$ can be verified:
\begin{equation}\label{eq:abstain-prop}
     \abstain{\alpha}(p) = \begin{cases}
     \argmax_{y \in \Y} p(y) & \max_y p(y) \geq 1 - \alpha\\
     \bot & \text{otherwise}
     \end{cases}~.~
\end{equation}

\cite{ramaswamy2018consistent} study the abstain property in depth, presenting a $\ceil {\log_2(n)}$ dimensional embedding of the abstain property.
However, it is unclear if this bound is tight, as the previously studied lower bounds of~\cite{ramaswamy2016convex} do not work well for this property, failing to give any lower bound tighter than the trivial $1$.

With our $1$-dimensional characterization, we already observe a tighter lower bound.
\begin{proposition}
	For $n \geq 3$ and $\alpha < 1$, the abstain loss $\ell_{\alpha}$ is not $1$-embeddable.
\end{proposition}
\begin{proof}
	Consider the intersection graph of $\gamma := \abstain{\alpha}$: the node associated with $\gamma_\bot$ has $n$ edges, and since we assume $n \geq 3$, it cannot be a path graph.
	For an example with $n=3$, see Figure~\ref{fig:abstain-alpha-half}.
\end{proof}

@@

\begin{proposition}
  The quadratic feasibility program (Definition \ref{def:qfp}) with input $C = \gamma_{\bot} = \{p \in \simplex: \max_y p(y) \leq 1/2\}$, $n=5$, and dimension parameter $d=2$, is infeasible.
\end{proposition}
\begin{corollary}
  The abstain loss with $\alpha=1/2$ on $5$ outcomes is not $2$-embeddable.
\end{corollary}

\bo{Am assuming that we will have to move almost everything to the appendix, wrote this blurb instead.}
We have two proofs, both delegated to the appendix.
The first is to directly obtain from mathematical software that the QFP is infeasible.
We use observations about $\ell_{1/2}$ to make a number of simplifications to the program first.
This reduces the QFP to the following problem:
\begin{quotation}
  Given a real-valued matrix $M$ in which some entries are missing, but have bounds of the form $[a,b]$, do there exist legal values for these entries such that $M$ is rank $d$?
\end{quotation}
The second proof is a direct geometrical one utilizing the same observations.
We show that packing too many normal vectors into a half-circle yields contradictions, implying that a $5$-outcome embedding into $2$ dimensions is impossible (and incidentally characterizing the structure of all possible $4$-dimensional embeddings).

\bo{Assuming everything after this will go to the appendix.}
The strategy is to consider the level set $\gamma_{\bot}$, the set of distributions with modal mass at most $1/2$.
We show that the quadratic feasibility program with this input cannot be satisfied with dimension $2$ for $n=5$.

\begin{lemma} \label{lemma:abstain-v-h}
  For the abstain loss $\ell_{1/2}$, the level set of abstain satisfies $\gamma_{\bot} = \conv\{(\delta_y + \delta_{y'})/2 : y,y' \in \Y, y < y'\} = \{p : Bp \geq \vec{0}\}$ where $\delta_y$ puts probability one on $y$ and $B = \ones \ones^{\intercal} - 2I \in \reals^{5 \times 5}$, i.e. has entries $-1$ on the diagonal and $1$ everywhere else.
\end{lemma}
\begin{proof}
  Recall that $\gamma_{\bot}$ is the set of distributions $p$ with $\max_y p(y) = 1/2$.
  First, note that each distribution of the form $(1/2, 1/2, 0, 0, 0)$ and so on is in $\gamma_{\bot}$.
  Meanwhile, every such $p$ can be written as a convex combination of these corners.
  Second, note that if $p \in \gamma_{\bot}$, then $p(y) \leq 1/2$ for all $y \in \Y$.
  These constraints can be rewritten as $\inprod{p}{b} \geq 0$ where $b_y = -1$ and $b_{y'} = 1$ for all $y' \neq y$, literally requiring $p(y) \leq \sum_{y' \neq y} p(y')$.\bo{Check notation $p_y$ versus $p(y)$.}
\end{proof}

\begin{observation} \label{obs:qfp-invariant-lin}
  For any invertible $A \in \reals^{d \times d}$, if $V,\{X^j : j \in [\ell]\}$ is a feasible solution to the quadratic feasibility program, then so is $(VA),\{A^{-1}X^j : j \in [\ell]\}$.
\end{observation}
\begin{proof}
  The halfspace constraints are $(VA)(A^{-1}X^j) \leq B \iff VX^{j} \leq B$.
  The $j^{th}$ vertex constraint is a vector equality $\sum_{y \in \Y} p^j(y) (A^{-1}X^j)_y = \vec{0}$.
  If we let $a_m$ be the $m$th row of $A^{-1}$, then the $m^{th}$ row of the vector equality is
  \begin{align*}
    0 &= \sum_{y \in \Y} p^j(y) \inprod{a_m}{x^j_y}  \\
      &= \inprod{a_m}{\sum_{y \in \Y} p^j(y) x^j_y}  \\
      &= 0
  \end{align*}
  so the program is feasible.
\end{proof}

\begin{corollary} \label{cor:qfp-wlog-v1}
  If there is a feasible solution to the quadratic feasibility program, then there is a feasible solution where $v_1$ is the first standard basis vector and $\|v_i\| \leq \|v_1\| = 1$ for all $i$.
\end{corollary}
\begin{proof}
  In particular, we can take a series of matrices $A$ in Observation \ref{obs:qfp-invariant-lin} that permutate the rows of $V$, scale\footnote{Note one can show $V = 0$ is not feasible unless $B$ is a trivial property, i.e. essentially has no rows at all.} all rows by $\frac{1}{\|v_1\|}$, and linearly map $v_1$ to $(1,0,\dots,0)$.
\end{proof}

@@

\begin{lemma} \label{lemma:vi-neq-cvj}
  There is no feasible solution to the QFP for $\gamma_{\bot}$ and $\ell_{1/2}$ where $v_i = c v_j$ for $c > 0$ and any $i \neq j$.
\end{lemma}
\begin{proof}
  There is an open halfspace through the origin strictly containing both the feasible regions $F_i = \{x : \inprod{v_i}{x} \leq -1, \inprod{v_j}{x} \leq 1 \forall j \neq i\}$ and $F_j$, so there is no set of witnesses such that $x^{ij}_i \in F_i$ and $x^{ij}_j \in F_j$, as this would contradict Lemma \ref{lemma:xiji-minus-xijj}.
\end{proof}

@@

\begin{figure}
	\begin{minipage}{0.3\linewidth}
		\centering
		\includegraphics[width=0.9\linewidth]{tikz/v1}
		\caption{Example of $v_1$.}
		\label{fig:v1}
	\end{minipage}
	\hfill
	\begin{minipage}{0.3\linewidth}
		\centering
		\includegraphics[width=0.9\linewidth]{tikz/qp-must-intersect-line-same-spot}
		\caption{If $x^{12}_1 \neq x^{13}_1$, we have a contradiction.}
		\label{fig:qp-must-intersect-line-same-spot}
	\end{minipage}
	\begin{minipage}{0.3\linewidth}
	\centering
	\includegraphics[width=0.9\linewidth]{tikz/qp-line-intersect-contradiction}
	\caption{The intersection of $\inprod{v_2}{x} = -1$ and $\inprod{v_3}{x} = 1$ occurs outside the region $\inprod{v_1}{x} \in [-1,1]$.}
	\label{fig:qp-line-intersect-contradiction}
\end{minipage}
\hfill

\end{figure}

@@

  If we consider the four lines $\inprod{v_2}{x}=1, \inprod{v_2}{x}=-1, \inprod{v_3}{x}=1, \inprod{v_3}{x}=-1$, we therefore have three points of intersection with the line $\inprod{v_1}{x}=1$ and three with the line $\inprod{v_1}{x}=-1$. 
  WLOG, these points from top left to top right are: $x^{12}_2$ (which equals $x^{13}_3$), the intersection with $\inprod{v_2}{x}=1$, and the intersection with $\inprod{v_3}{x}=1$; and therefore from bottom left to bottom right are: the intersection with $\inprod{v_3}{x}=-1$, the intersection with $\inprod{v_2}{x}=-1$, and the intersection with $x^{12}_1$ (which equals $x^{13}_1$).
  \bo{Needs more justification why these points are always in this order (up to renaming of $v_2$ and $v_3$).}

@@

\begin{proof}[Matrix rank proof]

@@

  \jessie{Fill in explanation} \bo{attempted}

@@

    \implies \inprod{v_i}{x^{ij}_i} = -1  \\

@@

    \implies \inprod{v_i}{x^{ji}_j} = 1  \\

@@

    \implies \inprod{v_i}{x^{jm}_j} \in [-1,1] .

@@

  \bo{I'm not sure the below is a proof, because you could have linear combinations with any coefficients, so I don't think this proves that the only solutions are negating rows. I think we have to appeal to Mathematica from here on out.}
  In characterizing the solutions to the smaller matrix rank problem, let us establish the linear equality of the first two rows and second two rows.
  (Note that any combination of row pairs works without loss of generality.)
  The first column containing $[-1 \; 1]^T$ implies that we must have the two rows being the negation of each other, and similarly for the third and fourth rows with the final column.
  This gives a matrix of rank 2 as follows:
  \[
    \begin{bmatrix}
  -1 & -1 & -1 & 1 & 1 & a \\
  1 & 1 & 1 & -1 & -1 & -a \\
  b & 1 & -1 & 1 & -1 & -1 \\
  -b & -1 & 1 & -1 & 1 & 1 \\
  \end{bmatrix}
  \] with $a, b \in [-1,1]$.
  One can verify that we cannot obtain a matrix of rank $1$, regardless of the choice of $a$ and $b$.
  Plugging in this solution set into the larger matrix yields
  \[
  \begin{bmatrix}
  {\color{red} -1} & {\color{red} -1} & {\color{red} -1} & {\color{red} 1} & {\color{red} 1} & {\color{red} a} & -1 & \cdot & \cdot & \cdot \\
  {\color{red} 1} & {\color{red} 1} & {\color{red} 1} & {\color{red} -1} & {\color{red} -1} & {\color{red} a} & \cdot & -1 & \cdot & \cdot \\
  {\color{red} b} & {\color{red} 1} & {\color{red} -1} & {\color{red} 1} & {\color{red} -1} & {\color{red} -1} & \cdot & \cdot & -1 & \cdot \\
  {\color{red} -b} & {\color{red} -1} & {\color{red} 1} & {\color{red} -1} & {\color{red} 1} & {\color{red} 1} & \cdot & \cdot & \cdot & -1 \\
  \cdot & \cdot & \cdot & \cdot & \cdot & \cdot & 1 & 1 & 1 & 1\\
  \end{bmatrix}
  \]

  In order for the $5^{th}$ row to be linearly dependent with the first two rows, we must have $r_5 = -r_1$ by the column $c_7$, and $r_5 = r_2$ by column $c_8$, yielding a contradiction.
  Similarly, we cannot establish linear independence with the last two rows because of restrictions imposed by columns $c_9$ and $c_{10}$.
  Therefore, the matrix completion problem given above has no solution of rank 3, and we conclude that $\abstain{1/2}$ is not $2$-embeddable with $n=5$.
\end{proof}

\section{Additional proofs}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: Abstain loss}\label{sec:examples}
\subsection{Abstain, $d=1$}\label{subsec:abstain-d1}
One classification-like problem that is of particular interest is the abstain property, elicited by the loss $\ellabs{\alpha}$ given in Equation~\eqref{eq:abstain}.
The property $\gamma = \abstain{\alpha}$ for $\alpha \in (0,1)$ can be verified:
\begin{equation}\label{eq:abstain-prop}
     \abstain{\alpha}(p) = \begin{cases}
     \argmax_{y \in \Y} p(y) & \max_y p(y) \geq 1 - \alpha\\
     \bot & \text{otherwise}
     \end{cases}~.~
\end{equation}

\cite{ramaswamy2018consistent} study the abstain property in depth, presenting a $\ceil {\log_2(n)}$ dimensional embedding of the abstain property.
However, it is unclear if this bound is tight, as the previously studied lower bounds of~\cite{ramaswamy2016convex} do not work well for this property, failing to give any lower bound tighter than the trivial dimension $1$.

With our $1$-dimensional characterization, we already observe a tighter lower bound.
\begin{proposition}
	For $n \geq 3$ and $\alpha < 1$, the abstain loss $\ell_{\alpha}$ is not $1$-embeddable.
\end{proposition}
\begin{proof}
	Consider the intersection graph of $\gamma := \abstain{\alpha}$: the node associated with $\gamma_\bot$ has $n$ edges, and since we assume $n \geq 3$, it cannot be a path graph.
	In fact, the incidence graph for this property is a star graph.
	For an example with $n=3$ and $\alpha = 1/2$, see Figure~\ref{fig:abstain-alpha-half}.
\end{proof}

@@

\begin{proposition}\label{prop:qfp-fails-abstain}
  The quadratic feasibility program (Definition \ref{def:qfp}) with input $C = \gamma_{\bot} = \{p \in \simplex: \max_y p(y) \leq 1/2\}$, $n=5$, and dimension parameter $d=2$, is infeasible.
\end{proposition}
\begin{corollary}
  The abstain loss with $\alpha=1/2$ on $5$ outcomes is not $2$-embeddable.
\end{corollary}

\bo{Am assuming that we will have to move almost everything to the appendix, wrote this blurb instead.}\jessie{Looks good to me.}
We have two proofs, both delegated to the appendix.
The first is to directly obtain from mathematical software that the QFP is infeasible.
We use observations about $\ell_{1/2}$ to make a number of simplifications to the program first.
This reduces the QFP to the following problem: 
\begin{quotation}
  Given a real-valued matrix $M$ in which some entries are missing, but have bounds of the form $[a,b]$, do there exist legal values for these entries such that $M$ is rank $d$?
\end{quotation}
The second proof is a direct geometric one utilizing the same observations.
We show that packing too many normal vectors into a half-circle yields contradictions, implying that a $5$-outcome embedding into $2$ dimensions is impossible (and incidentally characterizing the structure of all possible $4$-dimensional embeddings).

\section{$1$-dimensional characterization omitted proofs}

\section{$d$-dimensional omitted proofs}

\section{Simplifying the QFP for $\abstain{1/2}$, $d=2$}
In order to prove Proposition~\ref{prop:qfp-fails-abstain}, we take some simplifying steps to the quadratic feasibility program for this specific problem.
The strategy is to consider the level set $\gamma_{\bot}$, the set of distributions with modal mass at most $1/2$.
We show that the quadratic feasibility program with this input cannot be satisfied with dimension $2$ for $n=5$.

\begin{lemma} \label{lemma:abstain-v-h}
	For the abstain loss $\ell_{1/2}$, the level set of abstain satisfies $\gamma_{\bot} = \conv\{(\delta_y + \delta_{y'})/2 : y,y' \in \Y, y < y'\} = \{p : Bp \geq \vec{0}\}$ where $\delta_y$ puts probability one on $y$ and $B = \ones \ones^{\intercal} - 2I \in \reals^{5 \times 5}$, i.e. has entries $-1$ on the diagonal and $1$ everywhere else.
\end{lemma}
\begin{proof}
	Recall that $\gamma_{\bot}$ is the set of distributions $p$ with $\max_y p(y) = 1/2$.
	First, note that each distribution of the form $(1/2, 1/2, 0, 0, 0)$ and so on is in $\gamma_{\bot}$.
	Meanwhile, every such $p$ can be written as a convex combination of these corners.
	Second, note that if $p \in \gamma_{\bot}$, then $p(y) \leq 1/2$ for all $y \in \Y$.
	These constraints can be rewritten as $\inprod{p}{b} \geq 0$ where $b_y = -1$ and $b_{y'} = 1$ for all $y' \neq y$, literally requiring $p(y) \leq \sum_{y' \neq y} p(y')$.
\end{proof}

\begin{observation} \label{obs:qfp-invariant-lin}
	For any invertible $A \in \reals^{d \times d}$, if $V,\{X^j : j \in [\ell]\}$ is a feasible solution to the quadratic feasibility program, then so is $(VA),\{A^{-1}X^j : j \in [\ell]\}$.
\end{observation}
\begin{proof}
	The halfspace constraints are $(VA)(A^{-1}X^j) \leq B \iff VX^{j} \leq B$.
	The $j^{th}$ vertex constraint is a vector equality $\sum_{y \in \Y} p^j(y) (A^{-1}X^j)_y = \vec{0}$.
	If we let $a_m$ be the $m^{th}$ row of $A^{-1}$, then the $m^{th}$ row of the vector equality is
	\begin{align*}
	0 &= \sum_{y \in \Y} p^j(y) \inprod{a_m}{x^j_y}  \\
	&= \inprod{a_m}{\sum_{y \in \Y} p^j(y) x^j_y}  \\
	&= 0
	\end{align*}
	so the program is feasible.
\end{proof}

\begin{corollary} \label{cor:qfp-wlog-v1}
	If there is a feasible solution to the quadratic feasibility program, then there is a feasible solution where $v_1$ is the first standard basis vector and $\|v_i\| \leq \|v_1\| = 1$ for all $i$.
\end{corollary}
\begin{proof}
	In particular, we can take a series of matrices $A$ in Observation \ref{obs:qfp-invariant-lin} that permute the rows of $V$, scale\footnote{Note one can show $V = 0$ is not feasible unless $B$ is a trivial property, i.e. essentially has no rows at all.} all rows by $\frac{1}{\|v_1\|}$, and linearly map $v_1$ to $(1,0,\dots,0)$.
\end{proof}

@@

\begin{lemma} \label{lemma:vi-neq-cvj}
	There is no feasible solution to the QFP for $\gamma_{\bot}$ and $\ell_{1/2}$ where $v_i = c v_j$ for $c > 0$ and any $i \neq j$.
\end{lemma}
\begin{proof}
	There is an open halfspace through the origin strictly containing both the feasible regions $F_i = \{x : \inprod{v_i}{x} \leq -1, \inprod{v_j}{x} \leq 1 \; \forall j \neq i\}$ and $F_j$, so there is no set of witnesses such that $x^{ij}_i \in F_i$ and $x^{ij}_j \in F_j$, as this would contradict Lemma \ref{lemma:xiji-minus-xijj}.
\end{proof}

@@

\begin{figure}
	\begin{minipage}{0.45\linewidth}
		\centering
		\includegraphics[width=0.9\linewidth]{tikz/v1}
		\caption{Example of $v_1$.}
		\label{fig:v1}
	\end{minipage}
	\hfill
	\begin{minipage}{0.45\linewidth}
		\centering
		\includegraphics[width=0.9\linewidth]{tikz/qp-must-intersect-line-same-spot}
		\caption{If $x^{12}_1 \neq x^{13}_1$, we have a contradiction.}
		\label{fig:qp-must-intersect-line-same-spot}
	\end{minipage}
\end{figure}
\begin{figure}
		\begin{minipage}{0.4\linewidth}
		\centering
		\includegraphics[width=0.9\linewidth]{tikz/qp-intuition-contradiction.pdf}
		\caption{Reducing to the case where $x^{12}_1 = x^{13}_1$.}
		\label{fig:qp-line-intersect-contradiction}
	\end{minipage}
\hfill
	\begin{minipage}{0.59\linewidth}
	\centering
	\includegraphics[width=0.9\linewidth]{tikz/qp-intuition-contradiction-with-intersection.pdf}
	\caption{The intersection of $\inprod{v_2}{x} = -1$ and $\inprod{v_3}{x} = 1$ occurs outside the region $\inprod{v_1}{x} \in [-1,1]$.}
	\label{fig:qp-line-intersect-contradiction-with-intersection}
	\end{minipage}
\end{figure}

@@

	If we consider the four lines $\inprod{v_2}{x}=1, \inprod{v_2}{x}=-1, \inprod{v_3}{x}=1, \inprod{v_3}{x}=-1$, we therefore have three points of intersection with the line $\inprod{v_1}{x}=1$ and three with the line $\inprod{v_1}{x}=-1$. 
	WLOG, these points from top left to top right are: $x^{12}_2$ (which equals $x^{13}_3$), the intersection with $\inprod{v_2}{x}=1$, and the intersection with $\inprod{v_3}{x}=1$; and therefore from bottom left to bottom right are: the intersection with $\inprod{v_3}{x}=-1$, the intersection with $\inprod{v_2}{x}=-1$, and the intersection with $x^{12}_1$ (which equals $x^{13}_1$).
	\bo{Needs more justification why these points are always in this order (up to renaming of $v_2$ and $v_3$).}\jessie{Should we be able to take these to be in clockwise order WLOG in the grand scheme of trying to pack 5 normals into 180 degrees?  The order of the intersections follows from the clockwise order of the normals.  (ex: if the order was $v_2, v_1, v_3$, we would still get a contradiction, but in the first case like Figure~\ref{fig:intersection-graph-ex} I think)}

@@

\begin{proof}[Matrix rank proof of Proposition~\ref{prop:qfp-fails-abstain}]

@@

@@

	\implies \inprod{v_i}{x^{ij}_i} &= -1  \\

@@

	\implies \inprod{v_i}{x^{ji}_j} &= 1  \\

@@

	\implies \inprod{v_i}{x^{jm}_j} &\in [-1,1] .

@@

	\bo{I'm not sure the below is a proof, because you could have linear combinations with any coefficients, so I don't think this proves that the only solutions are negating rows. I think we have to appeal to Mathematica from here on out.}
	\jessie{Yeah definitely.  I hadn't touched this since before yesterday.}
	In characterizing the solutions to the smaller matrix rank problem, let us establish the linear equality of the first two rows and second two rows.
	(Note that any combination of row pairs works without loss of generality.)
	The first column containing $[-1 \; 1]^T$ implies that we must have the two rows being the negation of each other, and similarly for the third and fourth rows with the final column.
	\jessie{One can verify with mathematical software that $M_4$ has rank $2$ if it is in the following form:}
	\[
	\begin{bmatrix}
	-1 & -1 & -1 & 1 & 1 & a \\
	1 & 1 & 1 & -1 & -1 & -a \\
	b & 1 & -1 & 1 & -1 & -1 \\
	-b & -1 & 1 & -1 & 1 & 1 \\
	\end{bmatrix}
	\] with $a, b \in [-1,1]$.
	Plugging in this solution set into the larger matrix yields
	\[
	\begin{bmatrix}
	{\color{red} -1} & {\color{red} -1} & {\color{red} -1} & {\color{red} 1} & {\color{red} 1} & {\color{red} a} & -1 & \cdot & \cdot & \cdot \\
	{\color{red} 1} & {\color{red} 1} & {\color{red} 1} & {\color{red} -1} & {\color{red} -1} & {\color{red} a} & \cdot & -1 & \cdot & \cdot \\
	{\color{red} b} & {\color{red} 1} & {\color{red} -1} & {\color{red} 1} & {\color{red} -1} & {\color{red} -1} & \cdot & \cdot & -1 & \cdot \\
	{\color{red} -b} & {\color{red} -1} & {\color{red} 1} & {\color{red} -1} & {\color{red} 1} & {\color{red} 1} & \cdot & \cdot & \cdot & -1 \\
	\cdot & \cdot & \cdot & \cdot & \cdot & \cdot & 1 & 1 & 1 & 1\\
	\end{bmatrix}
	\]

	\jessie{Take this out?}
	In order for the $5^{th}$ row to be linearly dependent with the first two rows, we must have $r_5 = -r_1$ by the column $c_7$, and $r_5 = r_2$ by column $c_8$, yielding a contradiction.
	Similarly, we cannot establish linear independence with the last two rows because of restrictions imposed by columns $c_9$ and $c_{10}$.
	Therefore, the matrix completion problem given above has no solution of rank 3, and we conclude that $\abstain{1/2}$ is not $2$-embeddable with $n=5$.
\end{proof}
